---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  message = FALSE,
  warning = FALSE
)
```

# **Social Contact Matrices for Sub‑Saharan Africa**

[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.19174136.svg)](https://doi.org/10.5281/zenodo.19174136)

`ssamixr` provides harmonized, ready‑to‑use **social contact matrices** from empirical studies conducted across Sub‑Saharan Africa. The package includes **171 matrices** covering multiple countries, settings, and study designs, all standardized for infectious disease modeling and comparative epidemiology.

---

# 🚀 Installation

```{r, echo=TRUE}
# Install pak if needed
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}

# Install ssamixr package
pak::pak("fbkengne/ssamixr")

# Load the package
library(ssamixr)
```

🧭 Getting Started: A Simple Workflow  
Most users will want to:  

1. List all available matrices  
2. Filter matrices by country, study, or location  
3. Select a matrix ID  
4. Retrieve the matrix  
5. Visualize it  

1️⃣ List all matrices  

```{r, echo=TRUE}
all_mats <- list_matrices()
head(all_mats)
```

2️⃣ Filter matrices  
Example: matrices from Kenya

```{r, echo=TRUE}
kenya <- filter_matrices(country = "Kenya")
```

Example: household (HH) matrices  

```{r, echo=TRUE}
hh <- filter_matrices(location_type = "HH")
```

Example: combine filters  

```{r, echo=TRUE}
kenya_hh <- filter_matrices(country = "Kenya", location_type = "HH")
```

3️⃣ Select a matrix ID

```{r, echo=TRUE}
id <- kenya$matrix_id[1]
```

4️⃣ Retrieve the matrix

```{r, echo=TRUE}
m <- get_matrix(id)
```

5️⃣ Visualize the matrix
Default heatmap

```{r, echo=TRUE}
plot_matrix(m)
```

Gradient palette

```{r, echo=TRUE}
plot_matrix_gradient(m)
```

🔧 Visualization Parameters  
Both plot_matrix() and plot_matrix_gradient() support flexible, publication‑ready visualization of social contact matrices.  

Automatic label rotation  
By default:

If the matrix has more than 10 age groups, column labels rotate vertically

Otherwise, they remain horizontal

You can override this behavior:

```{r, echo=TRUE}
cameroon <- filter_matrices(country = "Cameroon")
id <- cameroon$matrix_id[1]
m <- get_matrix(id)
plot_matrix(m, col_label_orientation = "normal")
```

📐 Full Parameter Reference  
Parameter	Default	Description	Options  
m	—	A matrix returned by get_matrix().  
blank_color,	default = "grey90",	Fill color for missing values,	options(Any color).  
low_color (plot_matrix only),	default = "white",	Low end of gradient,	options(Any color).  
high_color (plot_matrix only),	default = "red",	High end of gradient,	options(Any color).  
value_text,	default = TRUE,	Print numeric values inside cells,	options(TRUE, FALSE).  
size,	default = 4.5	Text size for cell values,	options(Any positive number).  
col_label_orientation,	default = "auto",	Orientation of column labels,	options("auto", "normal", "vertical").  
base_size,	default = 16,	Base font size for the plot,	options(Any positive number0.  

🎨 Gradient Version  
plot_matrix_gradient() uses a fixed 4‑color palette:  

```{r, echo=TRUE}
c("#FFFFCC", "#9E9AC8", "#6A51A3", "#3F007D")
```

Everything else works the same as plot_matrix().

🧪 Example with custom options  

```{r, echo=TRUE}
plot_matrix(
  m,
  size = 5,
  col_label_orientation = "vertical",
  blank_color = "grey85"
)
```

📊 Dataset Summary   
The package contains:

- 171 social contact matrices
- covering 18 Sub‑Saharan African countries
- In Multiple settings:
  . All locations (All)
  . Household (HH)
  . School
  . Work
  . Other settings (Other)

- Each matrix includes:  

  . A numeric contact matrix  
  . Age‑group labels  
  . Study metadata  
  . Location type  
  . Country and author‑year identifiers  

🌍 Countries Included:  
The package includes matrices from the following 18 countries:   

- Cameroon
- Democratic Republic of Congo
- Ethiopia
- Gambia
- Ghana
- Guinea
- Ivory Coast
- Kenya
- Liberia
- Malawi
- Mozambique
- Nigeria
- Senegal
- Somalia/Somaliland
- South Africa
- Uganda
- Zambia
- Zimbabwe

(If you want, you can auto‑generate this list directly from the metadata.)

🏷️ Filtering Options  
Users can filter matrices using the following metadata fields:

Field	Description  
country:	Country name (18 total)  
study_id:	Unique study identifier  
author_year:	Author + publication year label  
location_type:	Contact setting: All, HH, School, Work, Other  

Example:

```{r, , echo=TRUE}
filter_matrices(
  country = c("Kenya", "Uganda"),
  location_type = "School"
)
```

📦 Overview  
ssamixr provides:  

Age‑structured social contact matrices for Sub‑Saharan African countries
Harmonized metadata for filtering and selection
Tools for loading, inspecting, and visualizing matrices

A reproducible workflow for integrating matrices into transmission models
Designed for:
- Infectious disease modelers
- Public health researchers
- Policy analysts
- Students learning age‑structured modeling

🌍 Data Sources  
The matrices included in ssamixr are derived from:
- Empirical contact surveys conducted across Sub‑Saharan Africa
- Harmonized demographic data
- Standardized processing pipelines ensuring comparability

Full details are available in the package vignette:

```{r, , echo=TRUE}
vignette("ssamixr")
```

📁 Package Structure  
ssamixr/  
├── R/                 # Functions  
├── data/              # Internal datasets  
├── inst/              # Metadata and documentation  
├── vignettes/         # Long-form documentation  
└── dev/               # Development scripts  

🤝 Contributing  
Contributions are welcome.  
If you would like to:  
- Add new matrices
- Improve documentation
- Report issues
- Suggest enhancements

please open an issue or submit a pull request on GitHub.

📄 License  
This package is released under the MIT License.
See LICENSE for details.

📄 Appendix: Full Social Contact Matrix Catalogue  
A complete appendix summarizing all 171 social contact matrices across 18 Sub‑Saharan African countries, including:  
- matrix counts by country  
- location types (household, school, work, other)  
- rural/urban distribution  
- COVID‑19 vs pre‑COVID periods  
- heatmaps and descriptions for each country  

The appendix is available in the package at:

```{r, , echo=TRUE}
system.file("extdata", "social_contact_appendix.pdf", package = "ssamixr")
```

📬 Citation  
If you use ssamixr in your research, please cite:  

Kengne FB, et al. (2026). ssamixr: Social Contact Matrices for Sub‑Saharan Africa.
A full citation entry is available via:  

```{r, echo=TRUE}
citation("ssamixr")
```

🧪 Reproducibility  
This package follows best practices for reproducible research:  

- Version‑controlled development  
- Documented data processing pipelines  
- Vignettes demonstrating usage  
- Stable GitHub releases with DOIs (via Zenodo)  

🙏 Acknowledgments  
We thank the researchers, survey teams, and collaborators who contributed to the collection and harmonization of social contact data across Sub‑Saharan Africa.
