---
title: "Introduction to ssamixr"
author: "Francis Barnabe Kengne et al. (2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ssamixr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ssamixr)
library(dplyr)
library(ggplot2)
```

# Load internal package data directly from the namespace

```{r}
data("country_list", package = "ssamixr")
data("ssa_matrices_meta", package = "ssamixr")
```

Overview
========

The ssamixr package provides access to 171 empirical social contact matrices
collected from a systematic review of studies conducted in Sub-Saharan Africa
(up to May 30, 2025). These matrices describe age‑specific contact patterns
useful for infectious disease modeling, public health analysis, and
epidemiological research.

This vignette introduces the core functionality of the package:

-Listing all available matrices

-Filtering matrices by country, study, author, or location type

-Retrieving a specific matrix

-Plotting matrices using built‑in heatmap utilities

Listing matrices

```{r}
list_matrices() %>% head()
```

Filtering matrices

Filter by country:

```{r}
filter_matrices(country = "Kenya")
```

Filter by location type:

```{r}
filter_matrices(location_type = "HH")
```

Retrieving a matrix

```{r}
m <- get_matrix("201")
m$matrix
m$row_labels
m$col_labels
```

Plotting matrices

To plot all matrices for a country:

```{r}
df_long <- build_country_long("Cameroon")
plot_country_heatmaps(df_long, "Cameroon")
```

To use the purple–yellow gradient:

```{r}
plot_country_heatmaps_gradientn(df_long, "Cameroon")
```

Citation

If you use ssamixr in your research, please cite:

Kengne et al. (2026). Social contact matrices in Sub‑Saharan Africa.

A full citation entry is available via:

```{r}
citation("ssamixr")
```

Conclusion

The ssamixr package provides a unified, reproducible interface for accessing,
filtering, and visualizing empirical social contact matrices from Sub‑Saharan
Africa. It is designed to support modeling workflows, comparative studies, and
public health decision‑making.
